Transfer of Learned Opponent Models in Zero Sum Games

### 2020-01-27 Report/draft_report_v1.Rmd
modify plots in exp2
added naive tom and qls with transfer

# Introduction
Being able to transfer previously acquired knowledge to a new domain is one of the hallmarks of human intelligence. Humans are naturally endowed with the ability to extract relevant features from a situation, identify the presence of these features in a novel setting and use previously acquired knowledge to adapt to previously unseen challenges using acquired knowledge. More formally, Perkins (1992) defines transfer of learning as the application of skills, knowledge, and/or attitudes that were learned in one situation to another learning situation. This typically human skill has so far eluded modern AI agents. Deep neural networks for instance can do very well on image recognition tasks and can even reach super-human performance levels on video and strategic board games. Yet they struggle to learn as fast or as efficiently as humans do, and more importantly they have a very limited ability to generalize and transfer knowledge to new domains. @lake2017building argue that human learning transfer abilities take advantage of important cognitive building blocks such as an abstract representation of concepts underlying tasks and compositionally structured causal models of the environment.

One way to build abstract representations of the environment when the task involves interactions with others is to build a model of the opponent strategy that may inform what actions they are likely to take next. Once we learn about the opponent's strategy, we can use this knowledge to inform how to best act in novel situations involving a previously encountered opponent. This may lead to very efficient generalization of knowledge, even to situations that are dissimilar to the history of interaction, assuming what we have learned about the opponent is an abstract representation that is not too dependent on the environment of the initial interaction.

There is evidence that people learn models of the opponents when they play repeated economic games [@stahl1995players], engage in bilateral negotiations [@baarslag2016learning], or simply try to exploit a non random player in chance games such as Rock-Paper-Scissors [@de2012higher]. In this paper, we are specifically interested in the way in which people build and use models of their opponent to facilitate learning transfer, when engaged in situations involving an interaction with strategic considerations. These situations arise frequently such as in negotiations, auctions, strategic planning and all other domains in which theory of mind abilities (Premack & Woodruff, 1978) play a role in determining human behaviour.

In order to explore learning transfer in strategic settings, it is generally useful to study simple games as a model of more complex interactions. More specifically, we need a framework that allows the study of whether and how a player takes into consideration, over time, the impact of its current and future actions on the future actions of the opponent and the future cumulative rewards. Repeated games, in which players interact repeatedly with the same opponent and have the ability to learn about the opponent's strategies and preferences (Mertens, 1990) are particularly adapted to the task of opponent modelling.

Early literature on learning transfer in games has mostly focused on measuring the proportion of people who play normatively optimal (Nash Equilibria) or salient actions (e.g Risk Dominance) in later games, having had experience with a similar game environment previously. For instance, Ho et al. (1998) measure transfer as the proportion of players who choose the Nash Equilibrium in later p-beauty contest games, after training on similar games. They find there is no evidence of immediate transfer (Nash equilibrium play in the first round of the new game) but positive structural learning transfer as shown by the faster convergence to equilibrium play by experienced vs non experienced players. Camerer & Knez (2000) test learning transfer in players exposed to two games with multiple equilibria sequentially and explore the ability of players to coordinate their actions to choose a particular equilibrium in subsequent games having reached it in prior ones. They distinguish between games that are similar in a purely descriptive way, meaning similar choice labels, identity of players, format and number of action choices; and games that are similar in a strategic sense, meaning similar payoffs from combination of actions, identical equilibrium properties or significant social characteristics of payoffs such as possibility of punishment, need for fairness and cooperative vs competitive settings. They find that transfer of learning (successful coordination) occurs more readily in the presence of both descriptive and strategic similarity. If the games were only strategically similar, then the transfer was much weaker.

Juvina et al.(2014)  made a similar distinction between what they deemed surface and deep similarities and find that both contribute to positive learning transfer. However, they show that surface similarity is not necessary for deep transfer and can either aid or block this type of transfer depending on whether it leads to congruent or incongruent actions in later games. In a series of experiments using economic signalling games, Cooper & Kagel (2003, 2004, 2008) find that participants who have learned to play according to Nash Equilibrium in one game can transfer this to subsequent games, even though the actions consistent with Nash Equilibrium in later games are different.  They show that this transfer is driven by the emergence of sophisticated players who are able to represent the strategic implications of their actions and reason about the consequences of changed opponent payoffs.

Most of these studies fail to offer a formal explanation of this transfer or a modelling framework that can explain the experimental observation of transfer between games and generalise it to extensive classes of games. A notable exception is the effort by Haruvy and Stahl (2012) to specify a model of learning where players learn abstract rules that they can generalise and transfer across dissimilar games, rather than action choices that can only be used within the same game. Participants played ten games, presented as 4x4 normal form (matrix payoffs). Their results suggest that subjects do transfer learning over descriptively similar but strategically dissimilar games and that this learning transfer is significant. They also showed that players learn abstract aspects of the game that are then transferred to new settings. Their rule-learning model, based on Stahl (1996), was able to capture participants' dynamic behavior and show that the propensity to select particular rules is perfectly transferred across games.

Still, one of the commonalities in studies of how humans adapt to computerised opponents is that they have mostly looked at the ability of players to detect and exploit action-based learning rules. The strategies implemented by the computer opponents had a style of play that was not "human-like" in the sense that humans are not very good at playing specific mixed strategies with any precision of at detecting patterns from long sequences of past play due to cognitive constraints. It is therefore important to have agents that "play like humans", and one way of achieving that is to embed theses agents with human-like theory of mind abilities based on limited steps of recursive reasoning. Simon (1972) explains that humans have limited cognitive capacities and as such cannot be expected to solve computationally intractable problems such as finding Nash equilibria. Instead, they will try to 'satisfice' by choosing a strategy that is adequate in a simplified model of the environment, rather than an optimal one. This concept finds its natural application in 'level-k' theory, first adopted by Stahl & Wilson (1995). It posits that deviations from Nash equilibrium solutions in simple games are explained by the fact that humans have a heterogenous degree of strategic sophistication. At the bottom of the ladder, level-0 players are non-strategic and play either randomly or use a salient strategy in the game environment (Arad & Rubinstein, 2012). Level-1 players are next up the ladder of strategic sophistication and will assume all their opponents belong to the level-0 category and as such will best respond to them given this assumption. Likewise, a level-2 player will choose actions that are the best response given the belief that all opponents are exactly one level below, and so on.

In this study, we propose to explore opponent modelling and its transfer with the use of computer agents possessing human-like theory of mind abilities with limited degrees of iterated reasoning. The agents will have a fixed strategy played stochastically and will either be level-1 or level-2 behavioral rule, mimicking human theory of mind abilities and the limited recursion depth they exhibit (Goodie et al., 2012).

Our choice of using computer oponents sinstead of matching groups of participants makes it easier  to disentangle the process of learning about the opponent from that of learning about the game structure and payoffs. When playing against other human opponents,players are learning about the game and the strategy of the opponent simultaneously. Thus, it is harder to focus on an individual and how her strategies are changing and adapting to the opponent's play if we cannot experimentally control the behaviour of the opponent. The use of computer opponents to elicit learning behavior has been explored in the literature with encouraging results.  For instance, Spiliopoulos (2013) made humans play constant sum games against 3 computer opponents programmed to take advantage of known patterns in human play such as imperfect randomization and heuristics use and found that human participants do adapt to the opponent they are facing. Shachat & Swarthout (2004) made human participants face computer opponents playing various mixed strategies in a zero-sum asymmetric matching pennies game. They found that the players changed their strategies towards exploiting the deviations from the Mixed Strategy Nash Equilibrium (MSNE), and that this exploitation was very likely if the deviation from the MSNE play was high.

@@

Next we compared the performance of players whose actions are consistent with each of our hypothesized models. Figure\ \@ref(fig:exp1-cumScores) shows the average cumulative performance of players across games, for participants grouped by which model best fits their behavior in experiment 1. We can see that participants whose actions are most consistent with learning an opponent model in a Bayesian way had the best overall performance (both with and witout transfer), followed by Q-learning conditional on last round play. EWA, QL and random players had, understandably the lowest performance.

```{r exp1-comp-model, fig.cap = "Histogram of best fitting computational models for experiment 1", fig.align='center'}
ggplot(data = dat_by_model, aes(x = t, y=model_acc_sum, group = Best_model)) + 
   geom_line(aes(color= Best_model))
```

@@

exp2_comp_results <- read.csv(file="../Experiment_2/exp2_results.csv")
exp2_comp_table <- table(exp2_comp_results[, "condition"],c("Random","ToM_BT","ToM_NBT", "ToM_NT", "QL_states","S_EWA","MBM")[apply(exp2_comp_results[,c("Random_BIC","Btwn_TR_BIC","No_Btwn_Tr_BIC","No_Tr_BIC","QL_states_BIC","EWA_self_BIC","MBM2_BIC")],1,which.min)])

@@

#knitr::include_graphics("../Report/images/exp1_comp_models.png", dpi = 108)
par(las=2) # make label text perpendicular to axis
par(mar=c(5,8,4,2)) # increase y-axis margin.
barplot(exp2_comp_table,
        horiz =TRUE, # rotate barplot for better visibility 
        las=1, # change orientation x axis labels 
        cex.names=0.8, # text label size
        beside =TRUE,
        xlab="Number of participants",
        # legend = c("level-1 opp first","level-2 opp first"),
        legend = paste0(rownames(exp2_comp_table) , " faced first"),
        args.legend=list( # positioning of legend box 
        x=ncol(exp1_comp_table) + 3,
        y=max(colSums(exp1_comp_table)) -1,
        bty = "n")
 )

@@

# Add best fitting model per participant
exp2_cum_score <- merge(exp2_cum_score, model_comp_2[, c("human_id", "Best_model")], by="human_id")

temp2 <- exp2_cum_score[,c("t","acc_sum","Best_model","condition","part_num")]
dat_by_model_2 <- temp2 %>% group_by(Best_model,t) %>% 
  summarize(model_acc_sum = mean(acc_sum))

@@

ggplot(data = dat_by_model_2, aes(x = t, y=model_acc_sum, group = Best_model)) + 
   geom_line(aes(color= Best_model)) +
  scale_x_continuous(minor_breaks = seq(0 , 180, 10), breaks = seq(0, 180, 60))

@@

Plotting cumulative scores by best model for experiment 2, we see very similar results looking at Figure\ \@ref(fig:exp2-cumScores), in that participants whose behavior was best fit by a bayesian model of learning the oppponent strategy had the highest cumulative performance, followed by theose relying on reinforcement learning with a state space of last round play, and as expectedm random play as well as EWA were at the bottom of cumulative performance.

```{r exp2-cumScores-bis, fig.cap = "Average cumulative scores of participants in experiment 2 by best fitting model", fig.align='center'}
# 
# knitr::include_graphics("../Report/images/exp2_cumScores.png", dpi = 108)

@@

```{r create_r-references}
# r_refs(file = "OM_transfer.bib")

>>>>

# Introduction
Being able to transfer previously acquired knowledge to a new domain is one of the hallmarks of human intelligence. Humans are naturally endowed with the ability to extract relevant features from a situation, identify the presence of these features in a novel setting and use previously acquired knowledge to adapt to previously unseen challenges using acquired knowledge. More formally, Perkins (1992) defines transfer of learning as the application of skills, knowledge, and/or attitudes that were learned in one situation to another learning situation. This typically human skill has so far eluded modern AI agents. Deep neural networks for instance can do very well on image recognition tasks and can even reach super-human performance levels on video and strategic board games. Yet they struggle to learn as fast or as efficiently as humans do, and more importantly they have a very limited ability to generalize and transfer knowledge to new domains. @Lake2017 argue that human learning transfer abilities take advantage of important cognitive building blocks such as an abstract representation of concepts underlying tasks and compositionally structured causal models of the environment.

One way to build abstract representations of the environment when the task involves interactions with others is to build a model of the opponent strategy that may inform what actions they are likely to take next. Once we learn about the opponent's strategy, we can use this knowledge to inform how to best act in novel situations involving a previously encountered opponent. This may lead to very efficient generalization of knowledge, even to situations that are dissimilar to the history of interaction, assuming what we have learned about the opponent is an abstract representation that is not too dependent on the environment of the initial interaction. There is evidence that people learn models of the opponents when they play repeated economic games [@stahl1995players], engage in bilateral negotiations [@baarslag2016learning], or simply try to exploit a non random player in chance games such as Rock-Paper-Scissors [@de2012higher]. In this paper, we are specifically interested in the way in which people build and use models of their opponent to facilitate learning transfer, when engaged in situations involving an interaction with strategic considerations. These situations arise frequently such as in negotiations, auctions, strategic planning and all other domains in which theory of mind abilities [@Woodruff] play a role in determining human behaviour.

In order to explore learning transfer in strategic settings, it is generally useful to study simple games as a model of more complex interactions. More specifically, we need a framework that allows the study of whether and how a player takes into consideration, over time, the impact of its current and future actions on the future actions of the opponent and the future cumulative rewards. Repeated games, in which players interact repeatedly with the same opponent and have the ability to learn about the opponent's strategies and preferences [@mertens1990repeated] are particularly adapted to the task of opponent modelling.

Early literature on learning transfer in games has mostly focused on measuring the proportion of people who play normatively optimal (Nash Equilibria) or salient actions (e.g Risk Dominance) in later games, having had experience with a similar game environment previously. For instance, @ho1998iterated measure transfer as the proportion of players who choose the Nash Equilibrium in later p-beauty contest games, after training on similar games. They find there is no evidence of immediate transfer (Nash equilibrium play in the first round of the new game) but positive structural learning transfer as shown by the faster convergence to equilibrium play by experienced vs non experienced players. @knez2000increasing test learning transfer in players exposed to two games with multiple equilibria sequentially and explore the ability of players to coordinate their actions to choose a particular equilibrium in subsequent games having reached it in prior ones. They distinguish between games that are similar in a purely descriptive way, meaning similar choice labels, identity of players, format and number of action choices; and games that are similar in a strategic sense, meaning similar payoffs from combination of actions, identical equilibrium properties or significant social characteristics of payoffs such as possibility of punishment, need for fairness and cooperative vs competitive settings. They find that transfer of learning (successful coordination) occurs more readily in the presence of both descriptive and strategic similarity. If the games were only strategically similar, then the transfer was much weaker.

@Juvina2013 made a similar distinction between what they deemed surface and deep similarities and find that both contribute to positive learning transfer. However, they show that surface similarity is not necessary for deep transfer and can either aid or block this type of transfer depending on whether it leads to congruent or incongruent actions in later games. In a series of experiments using economic signalling games [@cooper2003lessons; @Cooper2008] the researchers found that participants who have learned to play according to Nash Equilibrium in one game can transfer this to subsequent games, even though the actions consistent with Nash Equilibrium in later games are different.  They show that this transfer is driven by the emergence of sophisticated players who are able to represent the strategic implications of their actions and reason about the consequences of changed opponent payoffs.

Most of these studies fail to offer a formal explanation of this transfer or a modelling framework that can explain the experimental observation of transfer between games and generalise it to extensive classes of games. A notable exception is the effort by @Haruvy2012 to specify a model of learning where players learn abstract rules that they can generalise and transfer across dissimilar games, rather than action choices that can only be used within the same game. Participants played ten games, presented as 4x4 normal form (matrix payoffs). Their results suggest that subjects do transfer learning over descriptively similar but strategically dissimilar games and that this learning transfer is significant. They also showed that players learn abstract aspects of the game that are then transferred to new settings. Their rule-learning model, based on @stahl1995players, was able to capture participants' dynamic behavior and show that the propensity to select particular rules is perfectly transferred across games.

Still, one of the commonalities in studies of how humans adapt to computerised opponents is that they have mostly looked at the ability of players to detect and exploit action-based learning rules. The strategies implemented by the computer opponents had a style of play that was not "human-like" in the sense that humans are not very good at playing specific mixed strategies with any precision of at detecting patterns from long sequences of past play due to cognitive constraints. It is therefore important to have agents that "play like humans", and one way of achieving that is to embed theses agents with human-like theory of mind abilities based on limited steps of recursive reasoning. @simon1972theories explains that humans have limited cognitive capacities and as such cannot be expected to solve computationally intractable problems such as finding Nash equilibria. Instead, they will try to 'satisfice' by choosing a strategy that is adequate in a simplified model of the environment, rather than an optimal one. This concept finds its natural application in 'level-k' theory, first adopted by @stahl1995players. It posits that deviations from Nash equilibrium solutions in simple games are explained by the fact that humans have a heterogenous degree of strategic sophistication. At the bottom of the ladder, level-0 players are non-strategic and play either randomly or use a salient strategy in the game environment @arad201211. Level-1 players are next up the ladder of strategic sophistication and will assume all their opponents belong to the level-0 category and as such will best respond to them given this assumption. Likewise, a level-2 player will choose actions that are the best response given the belief that all opponents are exactly one level below, and so on.

In this study, we propose to explore opponent modelling and its transfer with the use of computer agents possessing human-like theory of mind abilities with limited degrees of iterated reasoning. The agents will have a fixed strategy played stochastically and will either be level-1 or level-2 behavioral rule, mimicking human theory of mind abilities and the limited recursion depth they exhibit @goodie2012levels.

Our choice of using computer oponents sinstead of matching groups of participants makes it easier to disentangle the process of learning about the opponent from that of learning about the game structure and payoffs. When playing against other human opponents, players are learning about the game and the strategy of the opponent simultaneously. Thus, it is harder to focus on an individual and how her strategies are changing and adapting to the opponent's play if we cannot experimentally control the behaviour of the opponent. The use of computer opponents to elicit learning behavior has been explored in the literature with encouraging results.  For instance, @spiliopoulos2013strategic made humans play constant sum games against 3 computer opponents programmed to take advantage of known patterns in human play such as imperfect randomization and heuristics use and found that human participants do adapt to the opponent they are facing. @shachat2004we made human participants face computer opponents playing various mixed strategies in a zero-sum asymmetric matching pennies game. They found that the players changed their strategies towards exploiting the deviations from the Mixed Strategy Nash Equilibrium (MSNE), and that this exploitation was very likely if the deviation from the MSNE play was high.

@@

Next we compared the performance of players whose actions are consistent with each of our hypothesized models. Figure\ \@ref(fig:exp1-cumScores) shows the average cumulative performance of players across games, for participants grouped by which model best fits their behavior in experiment 1. We can see that participants whose actions are most consistent with learning a ToM opponent model in a Bayesian way had the best overall performance (both with and witout transfer), followed by Q-learning conditional on last round play. EWA, QL and random players had, understandably the lowest performance.

```{r exp1-cumScores, fig.cap = "Average cumulative scores of participants in experiment 1 by best fitting model", fig.align='center'}
ggplot(data = dat_by_model, aes(x = t, y=model_acc_sum, group = Best_model)) + 
   geom_line(aes(color= Best_model))
```

@@

exp2_comp_results <- read.csv(file="../Experiment_2/exp2_results.csv")
exp2_comp_table <- table(exp2_comp_results[, "condition"],c("Random","ToM_BT","ToM_NBT", "ToM_NT", "ToM_Naive","QL_states", "QLS_within","S_EWA","MBM")[apply(exp2_comp_results[,c("Random_BIC","Btwn_TR_BIC","No_Btwn_Tr_BIC","No_Tr_BIC","Naive_BIC","QL_states_BIC","QLS_within_Tr_BIC","EWA_self_BIC","MBM2_BIC")],1,which.min)])

@@

#knitr::include_graphics("../Report/images/exp1_comp_models.png", dpi = 108)
par(las=2) # make label text perpendicular to axis
par(mar=c(5,8,4,2)) # increase y-axis margin.
barplot(exp2_comp_table,
        horiz =TRUE, # rotate barplot for better visibility 
        las=1, # change orientation x axis labels 
        cex.names=0.8, # text label size
        beside =TRUE,
        xlab="Number of participants",
        # legend = c("level-1 opp first","level-2 opp first"),
        legend = paste0(rownames(exp2_comp_table) , " faced first"),
        args.legend=list( # positioning of legend box 
        x=ncol(exp2_comp_table) + 6,
        y=max(colSums(exp2_comp_table)) +1,
        bty = "n")
 )

@@

# Add best fitting model per participant
exp2_cum_score <- merge(exp2_cum_score, model_comp_2[, c("human_id", "Best_model_2")], by="human_id")

temp2 <- exp2_cum_score[,c("t","acc_sum","Best_model_2","condition","part_num")]
dat_by_model_2 <- temp2 %>% group_by(Best_model_2,t) %>% 
  summarize(avg_acc_sum = mean(acc_sum))

@@

ggplot(data = dat_by_model_2, aes(x = t, y=avg_acc_sum, group = Best_model_2)) + 
   geom_line(aes(color= Best_model_2)) +
  scale_x_continuous(minor_breaks = seq(0 , 180, 10), breaks = seq(0, 180, 60))

@@

Plotting cumulative scores by best model for experiment 2, we see very similar results looking at Figure\ \@ref(fig:exp2-cumScores), in that participants whose behavior was best fit by a ToM model of learning the oppponent strategy had the highest cumulative performance, followed by theose relying on reinforcement learning with a state space of last round play, and as expected random play as well as EWA were at the bottom of cumulative performance.

@@

```{r create_r-references}
r_refs(file = "Mendeley.bib")
